Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-08-25 paper/main.tex
fiber section

\subsubsection{Topological Structure: Base Space \dbase}
\label{sec:atct:fb:base}
We encode the topological structure of the data as the \textcolor{base}{base space} \dbase\ of the fiber bundle since the base space acts as indexing space where every point in the base space has a corresponding fiber space $\pi^{-1}(\dbasepoint) = \dfiber_{\dbasepoint}$. The base space \dbase\ is a topological space $(\dbase,\mathcal{T}_{\dbase})$, which means it  consists of the set of points ${\dbasepoint \in \dbase}$ and topology $\mathcal{T}_{\dbase}$ \cite{munkresElementsAlgebraicTopology1984}. A topology $\dbase \coloneqq (\dbase,\mathcal{T}_{\dbase})$ is an axiomatic way of defining a topological structure\cite{bradleyTopologyCategoricalApproach2020}  as a collection of subsets, called open sets \footnote{Open sets (open subsets) are a generalization of open intervals to n dimensional spaces. For example, an open ball is the set of points inside the ball and excludes points on the surface of the ball. \cite{weissteinOpenSet}}, of that structure.

@@

The topological structure of the data can be expressed programmatically by constructing data types that encapsulate the class of topological structure-.i.e point, line, plane, network, etc. We propose that these data types can be formally specified as objects of a category. 
\begin{definition} 
   An \textbf{category} $\mathcal{C}$ consists of the following \textit{data}: 
\begin{enumerate}
  \item \textit{object}: X
  \item \textit{identity morphism}: $X \xrightarrow {id_x} X$
  \item \textit{morphisms}: $X \xrightarrow{f} Y$ between objects X, Y in $\mathcal{C}$
  \item \textit{composite morphism}: given the morphisms $X \xrightarrow{f} Y$ and $Y \xrightarrow{g} Z$, there exists a morphism $X \xrightarrow{g\circ f} Z$
\end{enumerate}
The data must satisfy the following \textit{rules}
\begin{enumerate}
  \item \textit{unitality:} for every morphism $ X \xrightarrow{f} Y$, there exists identity morphisms $id_x, id_y$ such that $f \circ id_x = f = id_y \circ f$
  \item \textit{associativity:} if any three morphisms $f, g, h$ are composable, 
    \begin{equation*}
      \begin{tikzcd}
        X \arrow[r, "f"] \arrow[rrr, "h\circ(g\circ f) = (h\circ g)\circ f"', bend right, dashed] & Y  \arrow[r, "g"] & Z \arrow[r, "h"] & W
        \end{tikzcd}
  \end{equation*}
  then they are associative such that $h\circ(g\circ f) = (h \circ g) \circ f$  \cite{lawvere2009conceptual,riehlCategoryTheoryContext,maclaneCategoriesWorkingMathematician2013,fongInvitationAppliedCategory2019}. 
  \end{enumerate}
\end{definition}

We construct a base space category $\mathcal{\dbasec}$ to encapsulate the topological structure of the data. The category $\mathcal{\dbase}$ has open set objects $\opensetc$ and inclusion morphisms $\opensetc_i \xrightarrow{\iota} \opensetc_j$ such that $\opensetc_i \subseteq \opensetc_j$. The composability property expresses that inclusion is transitive, while associativity expresses that the inclusion functions can be curried in various equivalent groupings. By formally specifying the properties of the topological structure datatypes as $\mathcal{\dbase}$, we can express that these are the properties that are required as part of the implementation of the data type objects.

\subsubsection{Data Fields: Fiber Space \dfiber}
\label{sec:atct:fb:fiber}
As mentioned in \autoref{sec:related-work:equivariance}, equivariance is traditionally evaluated as the preservation of field structure, which is based on the field type. To have a rich language for expressing field type, we generalize Spivak's fiber bundle formulation of table \cite{spivakDatabasesAreCategories2010,spivakSimplicialDatabases2009}. 
\note{def 3.5 and 3.6 should maybe be pushed back to prose as 3.7 is the only one we rely on - it defines what tau returns}
\begin{definition} A \textbf{type specification} is the bundle map $\pi: \mathscr{U} \rightarrow \textbf{DT}$. The base space $\textbf{DT}$ is the set of data types, and the total space $\mathscr{U}$ is the set of domains of those data types. Given an arbitrary data type $T \in \textbf{DT}$, the preimage $\pi^{-1} \subset \mathscr{U}$ is the domain of $T$ and $x \in \pi^{-1}(T)$ is an object of type $T$.  
\end{definition}
For example, if $T=\texttt{int}$, then the image $\pi^{-1}(\texttt{int}) = \mathbb{Z} \subset \mathscr{U}$ is the set of all integers. The domain $\mathscr{U}_{{\sigma}}$ is also a topological space, meaning there is no restriction on the dimensionality of the field, thus allowing for the encoding of multi-dimensional fields in the same manner as single dimensional fields. Since many fields can have the same datatype, Spivak formally defines a mapping from field name to field data type, akin to a  data base schema \cite{ullmanFirstCourseDatabase2008}. 
\begin{definition} A \textbf{schema} consists of a pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type. 
\end{definition}
The function $\sigma$ is composed with $\pi$ such that the bundle  $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ associates a field name $c \in C$ with the corresponding data type domain. Spivak then defines data records as sections on the bundle defined by $\pi_{\sigma}$. 
\begin{definition} A \textbf{record} is a function $\delement: \fnames \rightarrow \mathscr{U}_{\sigma}$ and the set of records on $\pi_{\sigma}$ is denoted $\Gamma^{\pi}(\sigma)$. Records must return an object of type $\sigma(\fname) \in \textbf{DT}$ for each field $c \in C$.
\end{definition}
Spivak then constructs tables as sections $\dsection: \dbase \rightarrow \Gamma^{\pi}(\sigma)$ from a key space $\dbase$ to the set of records $\Gamma^{\pi}(\sigma)$ on the schema bundle. Generalizing Spivak's formulation of tables, we define the fiber space in this work as the set of record functions 
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. The construction of the fiber as the space of records provides a way to distinguish data fields with the same datatype while allowing that datatype to be defined as almost any arbitrary topological space.

As with the base space category $\mathcal{\dbase}$, we propose a fiber category $\mathcal{\dfiberc}$ to encapsulate the field types of the data. The fiber category $\mathcal{\dfiberc}$ is a monoidal category, meaning the category has a single object $\dfiberc$ of an arbitrary type. The morphisms on the fiber object are $\dfunctc \in Hom(\dfiberc, \dfiberc)$. The fiber category is equipped with a bifunctor $\otimes: \mathcal{\dfiber} \times \mathcal{\dfiber} \rightarrow \mathcal{\dfiber}$. The bifunctor provides a method for combining fibers, thereby allowing for fields that contain multityped values. For example, wind can be represented as two fields $\dfiber_{speed} \times \dfiber_{direction}$ or a composite fiber field $\dfiber_{speed} \otimes \dfiber_{direction} = \dfiber_{wind}$. The $\otimes$ encapsulates both the sets associated with each fiber $\mathbb{R} \times \mathbb{R} = \mathbb{R}^{2}$ and the morphisms associated with each functor $(\dfunct_{speed}, \dfunct_{direction}) = \dfunct_{wind}$. Defining the field schema as a monoidal category allows us to formally express the structure of the field, as defined by its sets and functions, at different levels of composition as needed by different visualization tasks.

>>>>

\subsubsection{Topological Structure: Base Space \dbase}
\label{sec:atct:fb:base}
We encode the topological structure of the data as the \textcolor{base}{base space} \dbase\ of the fiber bundle since the base space acts as indexing space where every point in the base space has a corresponding fiber space $\pi^{-1}(\dbasepoint) = \dfiber_{\dbasepoint}$. The \textcolor{base}{base space} \dbasec\ is a topological space $(\dbase,\mathcal{T}_{\dbase})$, which means it  consists of the set of points ${\dbasepoint \in \dbase}$ and topology $\mathcal{T}_{\dbase}$ \cite{munkresElementsAlgebraicTopology1984}. A topology $\dbase \coloneqq (\dbase,\mathcal{T}_{\dbase})$ is an axiomatic way of defining a topological structure\cite{bradleyTopologyCategoricalApproach2020}  as a collection of subsets, called open sets \footnote{Open sets (open subsets) are a generalization of open intervals to n dimensional spaces. For example, an open ball is the set of points inside the ball and excludes points on the surface of the ball. \cite{weissteinOpenSet}}, of that structure.

@@

The topological structure of the data can be expressed programmatically by constructing data types that encapsulate the class of topological structure-.i.e point, line, plane, network, etc. We propose that these data types can be formally specified as objects of a category. 
\begin{definition} 
   An \textbf{category} $\mathcal{C}$ consists of the following \textit{data}: 
\begin{enumerate}
  \item a collection of \textit{objects} $X \in \mathcal{Ob}(\mathcal{C})$
  \item for every pair of objects $X, Y \in \mathcal{Ob}(\mathcal{C})$, a set of \textit{morphisms} $X \xrightarrow{f} Y \in Hom_{\mathcal{C}}(X, Y)$
  \item for every object $X$, a distinct \textit{identity morphism} $X \xrightarrow {id_x} X$ in $Hom_{\mathcal{C}}(X, X)$
  \item a \textit{composition function} $f \in Hom_{\mathcal{C}}(X, Y) \times  g \in Hom_{\mathcal{C}}(Y, Z) \rightarrow g \circ f \in Hom_{\mathcal{C}}(X, Z)$
\end{enumerate}
such that 
\begin{enumerate}
  \item \textit{unitality:} for every morphism $ X \xrightarrow{f} Y$, $f \circ id_x = f = id_y \circ f$
  \item \textit{associativity:} if any three morphisms $f, g, h$ are composable, 
    \begin{equation*}
      \begin{tikzcd}
        X \arrow[r, "f"] \arrow[rrr, "h\circ(g\circ f) = (h\circ g)\circ f"', bend right, dashed] & Y  \arrow[r, "g"] & Z \arrow[r, "h"] & W
        \end{tikzcd}
  \end{equation*}
  then they are associative such that $h\circ(g\circ f) = (h \circ g) \circ f$  \cite{lawvere2009conceptual,riehlCategoryTheoryContext,maclaneCategoriesWorkingMathematician2013,fongInvitationAppliedCategory2019}. 
  \end{enumerate}
\end{definition}

We encapsulate the topological structure of the data as a category $\mathcal{\dbasec}$. The standard construction of a category from a topological space is that it has open set objects $\opensetc$ and inclusion morphisms $\opensetc_i \xrightarrow{\iota} \opensetc_j$ such that $\opensetc_i \subseteq \opensetc_j$\cite{riehlCategoryTheoryContext}. The composability property expresses that inclusion is transitive, while associativity expresses that the inclusion functions can be curried in various equivalent groupings. By formally specifying the properties of the topological structure datatypes as $\mathcal{\dbase}$, we can express that these are the properties that are required as part of the implementation of the data type objects.

\subsubsection{Data Fields: Fiber Space \dfiber}
\label{sec:atct:fb:fiber}
As mentioned in \autoref{sec:related-work:equivariance}, visualization researchers traditionally describe equivariance as the preservation of field structure, which is based on the field type. Spivak shows that data typing can be expressed in a categorical framework in his fiber bundle formulation of tables in relational databases \cite{spivakDatabasesAreCategories2010,spivakSimplicialDatabases2009}. In this work, we adopt Spivak's definitions of \textit{type specification}, \textit{schema}, \textit{record} because that allows us to use a dimension agnostic named typing system for the fields of our dataset that is consistent with the abstraction we are using to express the continuity. Spivak introduces a \textit{type specification} as a bundle map $\pi: \mathscr{U} \rightarrow \textbf{DT}$. The base space $\textbf{DT}$ is the set of data types, and the total space $\mathscr{U}$ is the disjoint union of the domains of those data types. Given an arbitrary data type $T \in \textbf{DT}$, the preimage $ \pi^{-1}(T) \subset \mathscr{U}$ is the domain of $T$ and $x \in \pi^{-1}(T)$ is an object of type $T$\cite{spivakSimplicialDatabases2009}. For example, if $T=\texttt{int}$, then the image $\pi^{-1}(\texttt{int}) = \mathbb{Z} \subset \mathscr{U}$ is the set of all integers. Since many fields can have the same datatype, Spivak formally defines a mapping from field name to field data type, akin to a database schema \cite{ullmanFirstCourseDatabase2008}. According to Spivak, a \textit{schema} consists of a pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type\cite{spivakSIMPLICIALDATABASESa}. The function $\sigma$ is composed with $\pi$ such that the bundle  $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ associates a field name $c \in C$ with the corresponding domain for that data type $\mathscr{\U}_{\sigma}$. 
\begin{definition} A \textbf{record} is a function $\delement: \fnames \rightarrow \mathscr{U}_{\sigma}$ and the set of records on $\pi_{\sigma}$ is denoted $\Gamma^{\pi}(\sigma)$. Records must return an object of type $\sigma(\fname) \in \textbf{DT}$ for each field $c \in C$.
\end{definition}
Spivak then describes tables as sections $\dsection: \dbase \rightarrow \Gamma^{\pi}(\sigma)$ from a key space $\dbase$ to the set of all possible records $\Gamma^{\pi}(\sigma)$ on the schema bundle, and his notion of a table generalizes to our notion of a data container.

To build on the rich typing system provided by Spivak, we define the \textcolor{fiber}{fiber space} \dfiberc\ to be the set of all data records
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. 
Adopting Spivak's fiber bundle construction of types allows our model to reuse types so long as the field names are distinct and that field values can be accesed by field name,  since those are sections on $\mathcal{U}_{\sigma}$. Furthermore, since domains $\mathscr{U}_{{\sigma}}$ of types are a topological space, multi-dimensional fields can be encoded in the same manner as single dimensional fields and fields can have different names but the same type.

As with the base space category $\mathcal{\dbase}$, we propose a fiber category $\mathcal{\dfiberc}$ to encapsulate the field types of the data. The fiber category has a single object $\dfiberc$ of an arbitrary type and morphisms on the fiber object $\dfunctc \in Hom(\dfiberc, \dfiberc)$. The fiber category $\mathcal{\dfiberc}$ is a monoidal category, meaning that it is a category equipped with a bifunctor $\otimes: \mathcal{\dfiber} \times \mathcal{\dfiber} \rightarrow \mathcal{\dfiber}$. The bifunctor provides a method for combining fibers, thereby allowing us to express fields that contain multityped values. For example, wind can be represented as two fields $\dfiber_{speed} \times \dfiber_{direction}$ or a composite fiber field $\dfiber_{speed} \otimes \dfiber_{direction} = \dfiber_{wind}$. The $\otimes$ encapsulates both the sets associated with each fiber $\mathbb{R} \times \mathbb{R} = \mathbb{R}^{2}$ and the morphisms associated with each functor $(\dfunct_{speed}, \dfunct_{direction}) = \dfunct_{wind}$. Defining the field schema as a monoidal category allows us to formally express the structure of the field, as defined by its sets and functions, at different levels of composition as needed by different visualization tasks.
